Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-04-01 linear_adia.tex
revising in progress by mkl
revising

Rapid thermal relaxation is required to overcome a stabilizing
vertical entropy gradient \citep{goldreich67,urpin98,urpin03}. 
Eq. \ref{iso_vsi_cond} quantifies this requirement for vertically
global disturbances (which have small $M$).  We can interpret 
$\beta_\mathrm{crit}$ as the thermal timescale below which 
perturbations are effectively isothermal and the VSI operates. We see
that in a thin disk, a very short thermal timescale is required
($\beta_\mathrm{crit}\ll 1$) because $\beta_\mathrm{crit}\propto
\epsilon$.

>>>>

%Rapid thermal relaxation is required to overcome a stabilizing
%vertical entropy gradient \citep{goldreich67,urpin98,urpin03}. 
%Eq. \ref{iso_vsi_cond} quantifies this requirement for vertically
%global disturbances (which have small $M$).

We can interpret  $\beta_\mathrm{crit}$ as the thermal timescale below
which buoyancy becomes ineffective and perturbations are effectively
isothermal so the VSI can operate.  We see that in a thin disk, a very short thermal timescale
is required ($\beta_\mathrm{crit}\ll 1$) because
$\beta_\mathrm{crit}\propto \epsilon$. Physically, this is because
vertical shear is weak in a thin disk, so even a small entropy
gradient can significantly reduce VSI growth rates.

### 2015-04-01 numerical.tex
revising, added soft intro
revising

Complications from vertical boundaries arise when considering large
$\khat$. For $\khat=30$ and $\zmax=5H,\,7H$ we find for $\beta\gtrsim
0.1$ the dominant disturbance correspond to the leftmost eigenvalues 
seen in the lower two panels of Fig. \ref{compare_modes_cool_kx30},
which do not exist for $\zmax=3H$. For $\khat=50,\,100$ and
$\beta\to0$, we find surface modes dominate so the maximum growth rate
depends on $\zmax$. However, the maximum growth rate is sensitive to
$\zmax$ for moderately large $\beta$ ($\gtrsim 0.1$), i.e. close to
the stabilization of the fundamental mode. 
%not surprising since we enable strongly stabilizing buoyancy forces
%at large zmax

Our result is consistent with numerical simulations performed by
\cite{nelson13} for which a thermal relaxation timescale $\beta\gtrsim
0.6$ stabilized the VSI; as Fig. \ref{bcrit_compare1} show that growth
rates for $\beta\gtrsim 0.6$ are small. To destablize the disk with
$\beta\gtrsim O(1)$, one might consider 
$\khat > 100$ to increase the maximum growth rate by enabling surface
modes, but in a realistic disk
very small scale disturbances will likely be subject to viscous decay
(see \S\ref{application}). Numerical simulations are also unlikely to
resolve such small scales.

>>>>

Complications from vertical boundaries arise when considering large
$\khat$. For $\khat=30$ and $\zmax=5H,\,7H$ we find for $\beta\gtrsim
0.1$ the dominant disturbance correspond to the leftmost eigenvalues 
seen in the lower two panels of Fig. \ref{compare_modes_cool_kx30},
which do not exist for $\zmax=3H$. For $\khat=50,\,100$ and
$\beta\to0$, we find surface modes dominate so the maximum growth rate
depends on $\zmax$. However, the maximum growth rate is insensitive to
$\zmax$ for moderately large $\beta$ ($\gtrsim 0.1$), i.e. close to
the stabilization of the fundamental mode. 
%not surprising since we enable strongly stabilizing buoyancy forces
%at large zmax

Our result is consistent with numerical simulations performed by
\cite{nelson13} for which a thermal relaxation timescale $\beta\gtrsim
0.6$ stabilized the VSI; as Fig. \ref{bcrit_compare1} show that growth
rates for $\beta\gtrsim 0.6$ are small, and are restricted to very
small radial lengthscales $\sim 10^{-2}H$, for which current numerical
simulations may not be able to resolve.
%  To destablize the disk with
% $\beta\gtrsim O(1)$, one might consider 
% $\khat > 100$ to increase the maximum growth rate by enabling surface
% modes, but in a realistic disk
% very small scale disturbances will likely be subject to viscous decay
% (see \S\ref{application}). 
%Numerical simulations are also unlikely to
%resolve such small scales.
